---
title: "Midterm Project"
author: "Manning Smith"
date: "11/6/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(survival)
library(ranger)
library(ggplot2)
library(ggfortify)
library("survminer")
library(gridExtra)
library(caret)
library(knitr)
library(corrplot)
library(ggcorrplot)
library(mice)
library(tidyr)
library(VIM)
library(naivebayes)
library(MASS)
set.seed(15)
```

# Part 1

## Import Data
### Data Dictionary of the dataset
- PID:    Patient ID
- Age:    in years
- Gender:   F/M, Female/Male
- Ethnicity:
  - HISPANIC OR LATINO
  - NOT HISPANIC OR LA
  - NOT REPORTED
- White:
  - 0:  White
  - 1:  Non-white
- LVEF:   Left ventricular ejection fraction
- A1c:    Average amount of glucose (sugar) in blood
- SBP:    Systolic blood pressure,  mm Hg
- DBP:    Diastolic blood pressure,  mm Hg
- BMI:    Body mass index $\frac{kg}{m^2}$
- Ischemic Status:
  - Ischemic
  - Non-Ischemic
- Reason of End the Study:
  - STUDY TERMINATED BY SPONSOR
  - DEATH
  - LOST TO FOLLOW-UP
  - PHYSICIAN DECISION
  - WITHDEAWAL BY SUBJECT
- Months of Follow-up:  In months
- Group:
  - Placebo
  - Triinhibitor
- GFR:    Glomerular filtration rate

```{r import_1, echo=TRUE}
p1Data <- read.csv("BINF620BigData-1.csv")

#View(p1Data)
```

```{r cleanFilter, echo=TRUE}
# Clean Raw
p1Data <- p1Data %>% rename(Event_type = `Reason.of.End.the.Study`)
p1Data <- p1Data %>% rename(Time = `Months.of.Follow.up`)

# Filter Data
filtered_p1Data <- p1Data %>%
  filter(Event_type %in% c("DEATH", "STUDY TERMINATED BY SPONSOR"))
```

```{r cleanData2, echo=TRUE}
p1Data$Event <- ifelse(p1Data$Event_type == "DEATH", 1, 0)
p1Data$Gender <- ifelse(p1Data$Gender == "M", 0, 
                      ifelse(p1Data$Gender == "F", 1, NA))
p1Data$Group <- factor(p1Data$Group, levels = c("Placebo", "Triinhibitor"))

filtered_p1Data$Event <- ifelse(filtered_p1Data$Event_type == "DEATH", 1, 0)
filtered_p1Data$Gender <- ifelse(filtered_p1Data$Gender == "M", 0,
                      ifelse(filtered_p1Data$Gender == "F", 1, NA))
filtered_p1Data$Group <- factor(filtered_p1Data$Group, levels = c("Placebo", "Triinhibitor"))
```

```{r exploreData1, echo=TRUE, fig.height=4, fig.width=5}
summary(p1Data)

p1 <- ggplot(p1Data, aes(x = as.factor(Gender))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Gender (0 = Male, 1 = Female)", y = "Frequency", title = "Gender Frequency") +
  theme_minimal()

p2 <- ggplot(p1Data, aes(y = Age)) +
  geom_boxplot(fill = "lightblue") +
  labs(y = "Age", title = "Boxplot of Age Distribution") +
  theme_classic() + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())

p3 <- ggplot(p1Data, aes(x = as.factor(Gender), y = Age)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Gender (0 = Male, 1 = Female)", y = "Age", title = "Age Distribution by Gender") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2, nrow = 1)

ggplot(p1Data, aes(x = as.factor(Event))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Gender (0 = No Death, 1 = Death)", y = "Frequency", title = "Death Frequency") +
  theme_minimal()
```

```{r exploreData2, echo=TRUE, fig.height=4, fig.width=7}
summary(p1Data)

ggplot(p1Data, aes(x = as.factor(Event_type))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Reason", y = "Frequency", title = "Reason for the End of the Study") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
Notice that there are only a few records for `Lost to Follow-up`, `Physician Decision`, and `Withdrawal by Subject`. Alongside of the generic test, another test will be done to see if there is a difference in results with these events filtered out.

The result of filtering out these events.

```{r km, echo=TRUE}
# Kaplan–Meier method
km_fit <- survfit(Surv(Time, Event) ~ Group, data=p1Data)
p1 <- autoplot(km_fit)

# Filtered
km_fit_fil <- survfit(Surv(Time, Event) ~ Group, data=filtered_p1Data)
p2 <- autoplot(km_fit_fil)

grid.arrange(p1, p2, ncol = 1, nrow = 2)
```

```{r km_compare, echo=TRUE}
km_fit
km_fit_fil
```

Notice two different survival analysis have been done.

`Surv(Time, Event) ~ Group`: where Time is `Months.of.Follow.up` and Event is `Reason.of.End.the.Study` categorized into a binary relationship as either Death or Other and Group is factored by Placebo or Treatment.

`Surv(Time, Event) ~ Group`: where Time is `Months.of.Follow.up` and Event is `Reason.of.End.the.Study` categorized into a binary relationship as Death or "STUDY TERMINATED BY SPONSOR" and Group is factored by Placebo or Treatment.

Note there was no difference is these analysis, thus we will continue with the complete unfiltered dataset.

```{r cox, echo=TRUE}
cox_fit <- coxph(Surv(Time, Event) ~ Group + Age + Gender + Ethnicity + White + 
                 LVEF + A1c + SBP + DBP + BMI + Ischemic.Status + GFR, data = p1Data)

summary(cox_fit)
```

```{r hazard_ratio, echo=TRUE}
haz_ratio <- exp(coef(cox_fit))  # Hazard Ratios
haz_ratioINT <- exp(confint(cox_fit))  # 95% CI for Hazard Ratios

# Schoenfeld residuals test for proportional hazards assumption
ph_test <- cox.zph(cox_fit)
ph_test
```



# Part 2

## Import Data
### Data Dictionary of the dataset
- ID:  Subject Identifier
- ALIVE:  1--infant alive; 2—infant not alive;
- BF5EVER: Breastfeed--ever, 1=No, 2=Yes.
- BF5LNGTH: duration -- length of time in weeks
- PGWT_GN: Pregnant weight gain/lost
- MOM_CM:  Moms height – centimeters
- MOM_BMI:  MOM BODY MASS INDEX
- MOMCIG:   Number of Cigarettes Per Day
- CIG_PRIOR : No. of cigarettes smoked - prior to pregnant 
- MOM_LBKG : Moms weight -- before pregnant
- DDS_INS DDS: insurance
  - .A = NOT APPLICABLE.B = DK/BLANK
  - .N = NOT RECORDED
  - .U = UNKNOWN
  - 1 = NO
  - 2 = YES
- MH_PPDPR MH: depress since birth
  - .B = BLANK/DK
  - .S = SKIP
  - 1 = ALWAYS
  - 2 = OFTEN/ALMOST ALWAYS
  - 3 = SOMETIMES
  - 4 = RARELY
  - 5 = NEVER

```{r import_2, echo=TRUE}
p2Data <- read.csv("PARMSOct2020.csv")

#View(p2Data)
```

```{r exploreData3, echo=TRUE}
summary(p2Data)
str(p2Data)
```

```{r DataPeak1, echo=TRUE, fig.height=4, fig.width=7}
p1 <- ggplot(p2Data, aes(x = as.factor(ALIVE))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Alive (1 = Alive, 2 = Dead)", y = "Frequency", title = "Alive Frequency") +
  theme_minimal()

p2 <- ggplot(p2Data, aes(y = MOM_LBKG)) +
  geom_boxplot(fill = "lightblue") +
  labs(y = "Mom Weight (Before Pregant)", title = "Mom Weight Before Pregant") +
  theme_minimal()

p3 <- ggplot(p2Data, aes(y = MOM_BMI)) +
  geom_boxplot(fill = "lightblue") +
  labs(y = "Mom BMI", title = "Mom BMI") +
  theme_minimal()

p4 <- ggplot(p2Data, aes(y = CIG_PRIOR)) +
  geom_boxplot(fill = "lightblue") +
  labs(y = "Cigarettes Smoked", title = "Cigarettes Smoked Before Pregant") +
  theme_minimal()


grid.arrange(p2, p3, ncol = 2, nrow = 1)
grid.arrange(p1, p4, ncol = 2, nrow = 1)
```

```{r DataPeak2, echo=TRUE, fig.height=4, fig.width=7}
ggplot(p2Data, aes(x = as.factor(MH_PPDPR))) +
  geom_bar(fill = "skyblue") +
  labs(x = "MH_PPDPR", y = "Frequency", title = "Depress Since Birth Frequency") +
  theme_minimal()
```

```{r zeroVar, echo=TRUE}
nzv <- nearZeroVar(p2Data, saveMetrics = TRUE)
nzv_vars <- rownames(nzv[nzv$nzv == TRUE, ])
nzv_vars
```

```{r correlation1, echo=TRUE, fig.height=3, fig.width=5}
p1 <- ggplot(p2Data, aes(x = MOM_LBKG, y = MOM_CM)) +
  geom_point(alpha = 0.1) +
  labs(x = "Mom Weight Before Pregant (lbs)", y = "Mom Height (cm)", title = "Mom Weight vs Mom Height") +
  geom_smooth(method = "lm", se = FALSE)

p2 <- ggplot(p2Data, aes(x = MOM_LBKG, y = MOM_BMI)) +
  geom_point(alpha = 0.1) +
  labs(x = "Mom Weight Before Pregant (lbs)", y = "Mom BMI", title = "Mom Weight vs Mom BMI") +
  geom_smooth(method = "lm", se = FALSE)

grid.arrange(p1, p2, ncol = 2, nrow = 1)
```

```{r corrMatrix, echo=TRUE}
# Calculate and visualize the correlation matrix
cor_matrix <- cor(p2Data %>% select_if(is.numeric), use = "complete.obs")  # selecting numeric vars only

# Visualize correlation matrix
ggcorrplot(cor_matrix, method = "circle", type = "lower", lab = TRUE, title = "Correlation Matrix")
```

```{r correlation, echo=TRUE, fig.height=3, fig.width=5}
cor_pears <- cor(p2Data[, -1], method = "pearson")

cor_spear <- cor(p2Data[, -1], method = "spearman")

highCorr <- caret::findCorrelation(abs(cor_spear), cutoff = 0.95)
```

```{r imputeData, echo=TRUE}
p2_imputed <- mice(p2Data, m=5, method = 'pmm' ,maxit=5)
p2_complete <- complete(p2_imputed, 1)
```

```{r saveComplete, eval=FALSE, fig.height=7, fig.width=7, include=FALSE}
write.csv(p2_complete, "p2Imputed_complete.csv")
```

```{r loadComplete, eval=FALSE, fig.height=7, fig.width=7, include=FALSE}
p2_complete <- read.csv("p2Imputed_complete.csv")
```

```{r correlationImputed, echo=TRUE, fig.height=3, fig.width=5}
p1 <- ggplot(p2_complete, aes(x = MOM_LBKG, y = MOM_CM)) +
  geom_point(alpha = 0.1) +
  labs(x = "Mom Weight Before Pregant (lbs)", y = "Mom Height (cm)", title = "Mom Weight vs Mom Height") +
  geom_smooth(method = "lm", se = FALSE)

p2 <- ggplot(p2_complete, aes(x = MOM_LBKG, y = MOM_BMI)) +
  geom_point(alpha = 0.1) +
  labs(x = "Mom Weight Before Pregant (lbs)", y = "Mom BMI", title = "Mom Weight vs Mom BMI") +
  geom_smooth(method = "lm", se = FALSE)

grid.arrange(p1, p2, ncol = 2, nrow = 1)

ggplot(p2_complete, aes(x = as.factor(MH_PPDPR))) +
  geom_bar(fill = "skyblue") +
  labs(x = "MH_PPDPR", y = "Frequency", title = "Imputed - Depress Since Birth Frequency") +
  theme_minimal()
```

**Model:** `MH_PPDPR ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS` or `MH_PPDPR ~ PGWT_GN, MOM_CM, MOM_BMI, MOMCIG , CIG_PRIOR , MOM_LBKG, DDS_INS`.


```{r imputedPeak, echo=TRUE}
caret::featurePlot(x = p2_complete[, c("PGWT_GN", "MOM_CM", "MOM_BMI", "MOMCIG", "CIG_PRIOR", "MOM_LBKG", "DDS_INS")],
            y = as.factor(p2_complete$MH_PPDPR),
            plot = "density")
```

```{r binaryVar, echo=TRUE}
# Filter Data
filtered_p2Data <- p2_complete %>%
  filter(MH_PPDPR %in% c(1,2,4,5))

p2_complete$Event <- ifelse(p2_complete$MH_PPDPR %in% c(1, 2, 3), 1, 0)

filtered_p2Data$Event <- ifelse(filtered_p2Data$MH_PPDPR %in% c(1, 2), 1, 0)

ggplot(p2_complete, aes(x = as.factor(Event))) +
  geom_bar(fill = "skyblue") +
  labs(x = "MH_PPDPR (0 = Minimally Depressed, 1 = Experienced Depression", y = "Frequency", title = "Imputed - Depress Since Birth Frequency") +
  theme_minimal()

ggplot(filtered_p2Data, aes(x = as.factor(Event))) +
  geom_bar(fill = "skyblue") +
  labs(x = "MH_PPDPR (0 = Minimally Depressed, 1 = Experienced Depression", y = "Frequency", title = "Filtered Imputed - Depress Since Birth Frequency") +
  theme_minimal()

ggplot(trainData, aes(x = as.factor(MH_PPDPR))) +
  geom_bar(fill = "skyblue") +
  labs(x = "MH_PPDPR (0 = Minimally Depressed, 1 = Experienced Depression", y = "Frequency", title = "Filtered Imputed - Depress Since Birth Frequency") +
  theme_minimal()
```

```{r spliceData, echo=TRUE}
# Factor Variable
p2_complete$Event <- as.factor(p2_complete$Event)

trainIndex <- createDataPartition(p2_complete$Event, p = 0.8, list = FALSE)

trainData <- p2_complete[trainIndex, ]
testData <- p2_complete[-trainIndex, ]

trainData$Event <- as.factor(trainData$Event)
```

Baseline Logistic Regression Model


```{r logReg, echo=TRUE}
logit_model <- train(Event ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS,
                     data = trainData, method = "glm", family = binomial)
logit_model
```

```{r logReg_filtered, echo=TRUE}
filtered_p2Data$Event <- as.factor(filtered_p2Data$Event)

logit_model_filt <- train(Event ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS,
                     data = filtered_p2Data, method = "glm", family = binomial)
logit_model_filt
```

```{r naiveBayes, echo=TRUE}
naive_model <- naive_bayes(as.factor(Event) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
summary(naive_model)

naive_model_filt <- naive_bayes(as.factor(Event) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = filtered_p2Data)
summary(naive_model_filt)
```

```{r naiveBayes, echo=TRUE}
naive_model <- naive_bayes(as.factor(MH_PPDPR) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
summary(naive_model)
```

```{r predictAc, echo=TRUE}
# Predict
data_trn_pred <- predict(event_lda, trainData)
data_tst_pred <- predict(event_lda, testData)

# Store Predictions
predictions_trn <- data_trn_pred$class
predictions_tst <- data_tst_pred$class

# Calculate classification error
calc_class_err <- function(actual, predicted) {
  mean(actual != predicted)
}

# Check accuracy on the training set
accuracy_trn <- mean(predictions_trn == trainData$Event)

# Check accuracy on the test set
accuracy_tst <- mean(predictions_tst == testData$Event)

# Create confusion matrices
conf_matrix_trn <- table(predictions_trn, trainData$Event)
conf_matrix_tst <- table(predictions_tst, testData$Event)

# Output results
accuracy_trn
accuracy_tst
kable(conf_matrix_trn, caption = "Coefficient Matrix of Train | LDA", format = "html")
kable(conf_matrix_tst, caption = "Coefficient Matrix of Test  | LDA", format = "html")
```

```{r lda_qda, eval=FALSE, include=FALSE}
# lda
MH_PPDPR_lda = lda(as.factor(MH_PPDPR) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
MH_PPDPR_lda

event_lda = lda(as.factor(Event) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
event_lda

# qda
MH_PPDPR_qda = qda(as.factor(MH_PPDPR) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
MH_PPDPR_qda

event_qda = qda(as.factor(Event) ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)
event_qda
```

```{r lda_qda, eval=TRUE,}
# Lasso (glmnet package supports regularization methods)
library(glmnet)

x <- model.matrix(MH_PPDPR ~ PGWT_GN + MOM_CM + MOM_BMI + MOMCIG + CIG_PRIOR + MOM_LBKG + DDS_INS, data = trainData)[, -1]
y <- trainData$MH_PPDPR

model_lasso <- cv.glmnet(x, y, alpha = 1, family = "multinomial")  # alpha = 1 for Lasso
print(model_lasso)
```
