---
title: "BINF620 - Midterm"
author: "Manning Smith"
date: "11/6/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(survival)
library(ranger)
library(ggplot2)
library(ggfortify)
library("survminer")
library(gridExtra)
library(caret)
library(knitr)
library(corrplot)
library(ggcorrplot)
library(mice)
library(tidyr)
library(VIM)
library(naivebayes)
library(MASS)
set.seed(15)
```

# Import Data
```{r import, echo=TRUE}
# Part 1
p1Data <- read.csv("BINF620BigData-1.csv")
#View(p1Data)

# Part 2
p2Data <- read.csv("PARMSOct2020.csv")
#View(p2Data)
```

### Data Dictionary | Data Set 1
- PID:    Patient ID
- Age:    in years
- Gender:   F/M, Female/Male
- Ethnicity:
  - HISPANIC OR LATINO
  - NOT HISPANIC OR LA
  - NOT REPORTED
- White:
  - 0:  White
  - 1:  Non-white
- LVEF:   Left ventricular ejection fraction
- A1c:    Average amount of glucose (sugar) in blood
- SBP:    Systolic blood pressure,  mm Hg
- DBP:    Diastolic blood pressure,  mm Hg
- BMI:    Body mass index $\frac{kg}{m^2}$
- Ischemic Status:
  - Ischemic
  - Non-Ischemic
- Reason of End the Study:
  - STUDY TERMINATED BY SPONSOR
  - DEATH
  - LOST TO FOLLOW-UP
  - PHYSICIAN DECISION
  - WITHDEAWAL BY SUBJECT
- Months of Follow-up:  In months
- Group:
  - Placebo
  - Triinhibitor
- GFR:    Glomerular filtration rate
  
### Data Dictionary | Data Set 2
- ID:  Subject Identifier
- ALIVE:  1--infant alive; 2—infant not alive;
- BF5EVER: Breastfeed--ever, 1=No, 2=Yes.
- BF5LNGTH: duration -- length of time in weeks
- PGWT_GN: Pregnant weight gain/lost
- MOM_CM:  Moms height – centimeters
- MOM_BMI:  MOM BODY MASS INDEX
- MOMCIG:   Number of Cigarettes Per Day
- CIG_PRIOR : No. of cigarettes smoked - prior to pregnant 
- MOM_LBKG : Moms weight -- before pregnant
- DDS_INS DDS: insurance
  - .A = NOT APPLICABLE.B = DK/BLANK
  - .N = NOT RECORDED
  - .U = UNKNOWN
  - 1 = NO
  - 2 = YES
- MH_PPDPR MH: depress since birth
  - .B = BLANK/DK
  - .S = SKIP
  - 1 = ALWAYS
  - 2 = OFTEN/ALMOST ALWAYS
  - 3 = SOMETIMES
  - 4 = RARELY
  - 5 = NEVER
  
\newpage
# Part 1
## Introduction

```{r p1_dataPeak, echo=TRUE}
summary(p1Data)
```

## Data Exploration

```{r p1_cleanFilter, echo=TRUE}
# Clean Raw
p1Data <- p1Data %>% rename(Event_type = `Reason.of.End.the.Study`)
p1Data <- p1Data %>% rename(Time = `Months.of.Follow.up`)

# Filter Data
filtered_p1Data <- p1Data %>%
  filter(Event_type %in% c("DEATH", "STUDY TERMINATED BY SPONSOR"))

p1Data$Event <- ifelse(p1Data$Event_type == "DEATH", 1, 0)
p1Data$Gender <- ifelse(p1Data$Gender == "M", 0, 
                      ifelse(p1Data$Gender == "F", 1, NA))
p1Data$Group <- factor(p1Data$Group, levels = c("Placebo", "Triinhibitor"))

filtered_p1Data$Event <- ifelse(filtered_p1Data$Event_type == "DEATH", 1, 0)
filtered_p1Data$Gender <- ifelse(filtered_p1Data$Gender == "M", 0,
                      ifelse(filtered_p1Data$Gender == "F", 1, NA))
filtered_p1Data$Group <- factor(filtered_p1Data$Group, levels = c("Placebo", "Triinhibitor"))
```

```{r p1_exploreData, echo=TRUE, fig.height=4, fig.width=5}
summary(p1Data)

genFreq <- ggplot(p1Data, aes(x = as.factor(Gender))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Gender (0 = Male, 1 = Female)", y = "Frequency", title = "Gender Frequency") +
  theme_minimal()

ageDis <- ggplot(p1Data, aes(y = Age)) +
  geom_boxplot(fill = "lightblue") +
  labs(y = "Age", title = "Boxplot of Age Distribution") +
  theme_classic() + 
  theme(axis.title.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())

grid.arrange(genFreq, ageDis, ncol = 2, nrow = 1)

deathFreq <- ggplot(p1Data, aes(x = as.factor(Event))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Gender (0 = No Death, 1 = Death)", y = "Frequency", title = "Death Frequency") +
  theme_minimal()

studyEnd <- ggplot(p1Data, aes(x = as.factor(Event_type))) +
  geom_bar(fill = "skyblue") +
  labs(x = "Reason", y = "Frequency", title = "Reason for the End of the Study") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

grid.arrange(deathFreq, studyEnd, ncol = 2, nrow = 1)
```

## Analysis

```{r p1_models, echo=TRUE}
### Kaplan–Meier method
km_fit <- survfit(Surv(Time, Event) ~ Group, data=p1Data)
p1 <- autoplot(km_fit)

# Filtered
km_fit_fil <- survfit(Surv(Time, Event) ~ Group, data=filtered_p1Data)
p2 <- autoplot(km_fit_fil)

grid.arrange(p1, p2, ncol = 1, nrow = 2)

### Cox
cox_fit <- coxph(Surv(Time, Event) ~ Group + Age + Gender + Ethnicity + White + 
                 LVEF + A1c + SBP + DBP + BMI + Ischemic.Status + GFR, data = p1Data)

### Hazard
haz_ratio <- exp(coef(cox_fit))  # Hazard Ratios
haz_ratioINT <- exp(confint(cox_fit))  # 95% CI for Hazard Ratios

# Schoenfeld residuals test for proportional hazards assumption
ph_test <- cox.zph(cox_fit)
```

## Results

```{r results, echo=TRUE}
# View Results
km_fit
km_fit_fil
summary(cox_fit)
ph_test
```

### Conclusion


\newpage
# Part 2
## Introduction

## Data Exploration

## Analysis

## Results

### Conclusion
