---
title: "Title"
subtitle: "BINF620"
author: "A Manning Smith"
date: "12/15/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(matrixStats)
```

```{r dataSetup, include=FALSE}
# Load Filtered Data
load("filtered_NSCH.RData")
currData <- filteredData
load("modelData_NSCH.RData")
```

# Title

# Abstract

# Background
With a Global Pandemic behind us and mental health concerns (MHC) on the rise, researchers have intensified their focus on understanding the various social and family impacts experienced by children and adolescents during and after this unprecedented period. Based on data from 2021-2022, reported by the CDC 10% of children ages 3-17 had current, diagnosed anxiety. Moreover, 4% of children ages 3-17 had current, diagnosed depression. The rates grow when you examine teen groups specifically, 21% of teens reported symptoms of anxiety in the past two weeks and 17% reported symptoms of depression in the past two weeks.

~~~ Figure 1

This paper aims to explore the various impacts of one's childhood in shaping the mental health outcomes of children and adolescents. Using data from the 2022 National Survey of Children’s Health (NSCH) and the application of different statistical methods and neural networks to explore interactions between our selected variables, to identify key predictors that could positively impact the mental health of an individual.




# Study Design
Building upon this foundation, we are offering a robust and data-driven perspective on MHC among youth, outlining the statistical methodologies used to investigate these relationships.

## Aims
The goal of this research it to utilized our identified predictors and their association depression and anxiety in children and adolescents, ages 6-17. Our analyses are based on the 2022 NSCH Survey data and do not include clinical diagnoses or longitudinal evidence.

Furthermore:
- To examine how different family dynamics, social factors, and childhood experiences impact MHC.
- To examine the severity associated to different MHC.
- To explore the performance of the neural networks in predicting MHC.

## Data Exploration & Population Selection
The 2022 NSCH Survey is completed by parents/caregivers of children and able youth, containing 900+ data points. For the purposes of this research roughly 50 variables were selected for further analysis with a subset of about 20 variables utilized in the subsequent models.

Our selected population is children and adolescents, ages 6-17; these records can easily be sorted by the variable `age3_22` divided into 3 groups 0-5: group 1, 6-11: group 2, and 11-17: group 3. For our purposed we have filtered the data on group 2 and 3, reflecting our research interest.

There is no filter removing any specific race, but further analysis is needed to understand how the different selected predictors may vary among race subsequently impact the prediction of our outcome variable.

There are a various number of filters filtering out missing values as described in the independent variables section. The resulting sample was comprised of $19,039$ observations, with a binomial distribution centered around a mean age of approximately $12$ years old. This age range represents a critical period for social and mental development and is particularly significant due to its heightened vulnerability to emotional instability.

## Split and Train Data
For the process of splitting the data into a test and training set we will utilize a 80/20 split. With a large dataset it is appropriate to utilize this split allowing sufficient data to train then test our methods.
```{r split, eval = FALSE, include = TRUE}
n <- nrow(modelData)
train_size <- floor(0.8 * n)
train_indices <- sample(seq_len(n), train_size)

train_data <- modelData[train_indices, ]
test_data <- modelData[-train_indices, ]
```
We will utilize `train_data` in all models and then `test_data` to test the results of our model.

With the selected population and train split decided we can work to utilize our different statistical methods to analyze the strength of evidence in the predictive power of our statistical methods.


# Statistical Analysis
To ensure the viability of the chosen model, an initial logistic regression was performed to evaluate the predictive strength of the selected variables. This foundation will contribute to the selection of a model for a neural network to learn the relationships between predictors and mental health outcomes.

## Response Variable - Mental Health Concern
The selected variables of interest are depression and anxiety. These variables are recorded in a question as "Has a doctor or other health care provider EVER told you that this child has?...Anxiety Problems? or ...Depression?". Followed by, "If yes, does this child CURRENTLY have the condition?" The final question asks for a level of severity provided by 1 = Mild, 2 = Moderate, 3 = Severe.

For our research focus, we utilized three different outcome variables encoded from different reporting variables for depression and anxiety.

- `MHealthConcern`, a binary variable encoded as:
$$
\begin{align*}
  0 &= \text{`K2Q33B` or `K2Q32B`} = 2 \text{('No')} \\
  1 &= \text{`K2Q33B` or `K2Q32B`} = 1 \text{('Yes')}
\end{align*}
$$

- `MHealthScore`, a numerical categorical variable encoded as:
$$
\begin{align*}
  0 &= \text{`K2Q33A` or `K2Q32A`} = 2, &\text{ No reported depression or anxiety} \\
  1 &= \text{`K2Q33A` or `K2Q32A`} = 1, &\text{ Reported depression or anxiety} \\
  2 &= \text{`K2Q33B` or `K2Q32B`} = 1, &\text{ Currently has depression or anxiety} \\
  3 &= \text{`K2Q33B` and `K2Q32B`} = 1, &\text{ Currently has depression and anxiety}
\end{align*}
$$
This variable is helpful to identify differences between individuals who once had a MHC vs those that may or may not currently have a MHC.

- `MHealthScorePlus`, a numerical score variable encoded as:
$$
\begin{align*}
  &= \text{'MHealthConcern'} + \text{'K2Q33C'} + \text{'K2Q32C'}
\end{align*}
$$
This variable takes into account the severity of the MHC. The max score for this variable is 9 being that an individual currently has depression and anxiety and the severity of their depression and anxiety are listed as Severe.

Our predictor variables will allow us to predict our main outcome variable being the binary outcome for individuals currently experiencing depression or anxiety. This multi-level approach ensures a comprehensive understanding of how various predictors contribute to depression and anxiety in children and adolescents.

## Independent Variables
```{r mutations eval = TRUE, include = TRUE}

```
There are various factors that can impacts ones mental health. Stress is a big factor that contributes to ones' mental health. Adverse childhood experiences, family dynamics, social environments, and neighborhood dynamics are all forms that stress can develop and contribute to mental health concerns. Indicators such as parental MHC, emotional support providers, adverse community and family events, and household experiences all contribute to perceived stress and MHC.

Our variables of interest are listed below and their detailed descriptions can be found in the dictionary.
```
"SC_AGE_YEARS", "HHCOUNT", "BORNUSA", "K8Q35", "ACE12", 
"PHYSACTIV", "age3_22", "sex_22", "MotherMH_22", 
"FatherMH_22", "ScreenTime_22", "ACEct11_22", "ACE4ctCom_22", 
"SC_RACE_R", "bully_22", "bullied_22", "AftSchAct_22", 
"EventPart_22", "mentor_22", "ShareIdeas_22", 
"ACE6ctHH_22", "NbhdSupp_22", "NbhdSafe_22", "FAMILY_R"
```
The selected variables are grouped into the following themes:
- Individual Information
  - "SC_AGE_YEARS", "sex_22", "SC_RACE_R"
  - These variables take into account specific age, sec, and race. All of these predictors will play different but crucial roles for diversifying the selected population.
- Family History
  - "BORNUSA"
    - This variable is a binary variable representing if the individual was born in the USA.
  - "MotherMH_22", "FatherMH_22"
    - These variables signifies the severity of the mother/father mental health. This value is provided in:
      - 1: Excellent
      - 2: Good
      - 3: Fair or Poor
      - 95: No mother/father reported in the household as a primary caregiver of the child
      - 99: Missing Values.
```{r fig.height=4, fig.width=6, include=, mutation1 eval=TRUE}
long_data <- currData %>%
  pivot_longer(cols = c(MotherMH_22, FatherMH_22), 
               names_to = "Parent", 
               values_to = "Category") %>%
  group_by(Parent, Category) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(proportion = count / sum(count))

# Plot proportions
library(ggplot2)
ggplot(long_data, aes(x = factor(Category), y = proportion, fill = Parent)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Category",
    y = "Proportion",
    title = "Figure 0.0 | Proportions of Mental Health Severity by Categories of Parent",
    fill = "Parent"
  ) +
  scale_x_discrete(
    labels = c(
      "1" = "Excellent",
      "2" = "Good",
      "3" = "Fair or Poor",
      "95" = "No Primary Caregiver",
      "99" = "Missing Values"
    )
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "right") +
  scale_fill_brewer(palette = "Dark2")

```
For our research purposes we have removed all missing values from our dataset and have performed the following mutation to group the variable into binary representation. Due to this filter further analysis must be performed to account of the absence of parents in the child's life.
- Family Dynamic
  - "EventPart_22"
    - This variables expresses the value in parent participation in the child's life. The value is provided in:
      - 1 "Always"
      - 2 "Usually"
      - 3 "Sometimes"
      - 4 "Rarely or never"
      - 99 "Missing"
      - 90 "Children age 0-5 years"
This variable has been grouped into a binary representation and filtered out missing values.
  - "ACE6ctHH_22"
    - This variable is representative of the number of household-based adverse childhood experiences based on 6 household items listed from the survey. No filters were performed on this variable.
  - "FAMILY_R"
    - This variable categorizes the parent influence in the child's life. This value is provided in:
      - 1 = Two biogical/adoptive parents, currently married
      - 2 = Two biogical/adoptive parents, not currently married
      - 3 = Two parents (at least one not biological/adoptive), currently married
      - 4 = Two parents (at least one not biological/adoptive), not currently married
      - 5 = Single mother
      - 6 = Single father
      - 7 = Grandparent household
      - 8 = Other relation
  - "HHCOUNT"
    - This variable expresses how many people are living in the household.
- Neighborhood Situations / Experiences
  - "NbhdSafe_22"
    - This variable provides how safe the neighborhood is. The value is provided in:
      - 1 "Definitely agree"
      - 2 "Somewhat agree"
      - 3 "Somewhat/Definitely disagree"
This variable has been grouped into a binary representation and filter out missing values.
  - "ACE4ctCom_22"
    - This variable signifies if the child experienced 1 or more community based adverse events. The value is provided in:
      - 1 "No community-based adverse childhood experiences"
      - 2 "Experienced 1 or more community-based ACEs"
This variable did not filtered out missing values.
- Activities
  - "PHYSACTIV"
    - This variable represents how many days did this child exercise, play a sport, or participate in physical activity for at least 60 minutes. The value is provided in:
      - 1 = 0 days
      - 2 = 1 - 3 days
      - 3 = 4 - 6 days
      - 4 = Every day
  - "AftSchAct_22"
    - This variable represents if the child participates in organized activities outside school. It is a binary representation with missing values filtered out.
- Adverse Events
  - "ACE12"=
    - This varaible is a binary representation if the child experienced "Treated or judged unfairly because of their sexual orientation or gender identity". 0:No, 1: Yes
  - "ACEct11_22"
    - This variable represents the number of adverse events of the 11 possible.
- Social / Friends Interactions
  - "ScreenTime_22"
    - This variable represents the number of hours spent on a screen daily. Missing values have been filtered out.
  - "bully_22" and "bullied_22"
    - These varaibles represent if the child was bullied or is a bully. These values have filtered out missing values. The value is provided in:
      - 1 "Never (in the past 12 months)"
      - 2 "1-2 times (in the past 12 months)"
      - 3 "1-2 times per month"
      - 4 "1-2 times per week"
      - 5 "Almost every day" 
- Emotional Support
  - "K8Q35"
    - This variables signifies if the child has someone to turn too for emotional support. 0: No, 1: Yes.
  - "NbhdSupp_22"
    - This variable signifies the children who live in supportive neighborhood. 0: No, 1: Yes. Missing values have been filtered out.
  - "ShareIdeas_22"
    - This variable signifies how well children share ideas or talk about things that really matter with their parents. This variable has been filtered for missing values. The value is provided in:
      - 1 "Very Well"
      - 2 "Somewhat Well"
      - 3 "Not at all or Not very well"
  - mentor_22
  
## Initial Models
In alignment with our themes in the variables 4 different smaller models were formed to prove viability in proceeding with a neural network for the main model.

**Model 1:** ``

## Model Selection

## Neural Network 1
Neural network #1 with binary response of current mental health
## Neural Network 2
Neural network #2 with continuous response of score from past and current

# Results
## Model Used
## Model Evaluation Metrics
## Major Findings
## Tables and Figures

# Conclusions

# References