---
title: "Assignment 8"
author: "Manning Smith"
date: "12/10/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(rpart)
library(rpart.plot)
library(randomForest)
library(survival)
library(ranger)
library(ggplot2)
library(ggfortify)
library("survminer")
library(gridExtra)
library(caret)
library(mlbench)
library(knitr)
library(kableExtra)
library(corrplot)
library(ggcorrplot)
library(mice)
library(class)
library(tidyr)
library(VIM)
library(naivebayes)
library(MASS)
library(boot)
library(neuralnet)
library(GGally)
set.seed(43)
```

# Import Data
```{r import, echo=TRUE}
framData <- read.csv("Framingham.csv")
View(framData)
```

## Dictionary
- male:  male(1) or female (0) (Nominal)
- age:   age of the patient;( continuous)
- BPMeds:  the patient was on blood pressure medication or not (Nominal), “1”-- “Yes”, “0”-- “No”
- prevalentStroke: the patient had previously a stroke or not (Nominal), “1”-- “Yes”, “0”-- “No”
- prevalentHyp:  the patient was hypertensive or not (Nominal), “1”-- “Yes”, “0”-“No”
- diabetes:  patient had diabetes or not (Nominal), “1”-- “Yes”, “0”-- “No”
- totChol:  total cholesterol level (Continuous)
- sysBP:  systolic blood pressure (Continuous)
- diaBP:  diastolic blood pressure (Continuous)
- BMI:  Body Mass Index (Continuous)
- heartRate: heart rate (Continuous)
- glucose: glucose level (Continuous)

# Question 1

```{r Q1, echo=TRUE}
```

### 1a)

```{r Q1a, echo=TRUE}
```

\newpage
# Question 2

```{r Q2, echo=TRUE}
```

\newpage
# Question 3

```{r Q3, echo=TRUE}
```